Type Checker Generation using Reference Attribute Grammars

### 2024-01-26 MSc.tex
Add popsci summary
Address review comments

\thesisnumber{LU-CS-EX: 2023-79} % Magic Number! Do not change unless Birger Swahn asks you to do so!
% default is Master. Uncomment the following for "kandidatarbete"/Bachelor's thesis
%\thesistype{Bachelor}{Kandidatarbete}

@@

%\title{Formatting a Master's Thesis}
\title{Type Checker Generation using Reference Attribute Grammars}
\begin{otherlanguage}{swedish}
\svensktitel{Infoga den Svenska titeln här!}
\end{otherlanguage}

@@

\acknowledgements{
Sincere thanks to Christoph Reichenbach,
whose enthusiasm for the subject inspired me to pursue this subject
and whose guidance in matters of both implementation and writing were key to this thesis.
Additional thanks to Niklas Fors for providing several valuable insights, greatly simplifying certain components and Alexandru Dura for providing some guidance to JastAdd internals.

I would also like to thank my partner Lo-Isabel Krantz Andrée, my family, and Moritz for their support throughout the duration of this project.

%If you want to thank people, do it here, on a separate right-hand page. Both the U.S. \textit{acknowledgments} and the British \textit{acknowledgements} spellings are acceptable.
%
%We would like to thank Lennart Andersson for his feedback on this template.
%
%We would also like thank Camilla Lekebjer for her contribution on this template, as well as Magnus Hultin for his popular science summary class and example document.
%
%Thanks also go to the following (former) students for helping with feedback and suggestions on this template: Mikael Persson, Christoffer Lundgren, Mahmoud Nasser.
}

@@

We define a custom parsable language for typing rules adapted from natural semantics, and introduce a strategy to compile it into RAG attributes.
Our case studies show that simple type systems can be systematically compiled into attributes on the AST, and the tree syntax provides a natural approach for recursive evaluation.
Overall, this initial exploration shows that RAGs are a flexible approach for implementing type checkers for syntax-directed typing rules, though there are difficulties when implementing rule verification at compiler runtime.
%\CR{Usually you then put a high-level summary of the rest of the thesis here (not mentioning chapters yet), something like ``we examine X and Y and introduce a strategy for Z.
%  We evaluate the approach against Foo, and demonstrate Bar.
%  Our case studies show Quux.
%  Overall, this initial exploration shows \ldots''}

@@

\section{Abstract syntax trees}\label{abstractsyntaxtrees}
The syntax of a language is the definition of the symbols and structures which make up a valid program.
%\CR{What's a syntax?}
An abstract syntax tree (AST) is a simplified tree-based representation of the syntax.
Each node in the tree represents a syntactic construct of the language, such as expressions, statements or declarations and the branches represent the structure.
It is abstract in that it neglects to include all the concrete details of the syntax, but rather focuses on the structure of the program.
Structural elements of the source code, such as parenthesis, commas and brackets, are no longer necessary, as the structure of the tree conveys this semantic information.

This makes it easier to write analysis based on the behaviour of a program, without concerning oneself with its exact formulation.
Several different formulations of a program with identical behaviour may be abstracted into identical ASTs.
The hierarchical structure of the tree lends itself especially well for recursive analyses, and eases the passing of information between connected structures by connecting them as parent and children nodes.
%\CR{what is ``this information'', concretely?}
A statement like \verb|if (true) then {1} else {0}| may instead be represented by an if node, with named branches pointing towards the condition and the two values.
%\CR{You could show an example, but at the very least I would like to read an argument for why the hierarchical structure of a tree is helpful.}
%\CR{You can find common examples for showing the value of ASTs over concrete syntax trees in the EDAN65 slides.}

@@

\subsection{Type checking}
Verifying that these constraints are upheld can assist us in catching programming mistakes, preventing the wrong data types from being used in the wrong places, which may have led to erroneous or unpredictable behaviour.
There are two main approaches to ensuring the constraints are upheld, dynamic and static type-checking.
As our project generates a static type checker, we will focus our explanations on that topic, but provide a quick overview of their differences here.
%\CR{Why would we even want to ensure that they are upheld?}

@@

\subsubsection{Type syntax}\label{typesyntax}
We're using the term `type syntax' to mean a definition of the valid types within a language, often referred to simply as the set of types.
%\CR{Is ``type syntax'' what Pierce calls it?  Not sure I've heard that term before; I don't think it's wrong, personally, but please check in case he uses a different term.}
%\reviewquestion{I believe Pierce simply refers to it as the "set of types". I seem to have got the term type syntax from EDAP15 \url{https://fileadmin.cs.lth.se/cs/Education/EDAP15/2022/web/skills/type-system.html}. In my opinion it works better to have a clear name for it, so I'm keeping it for now, unless you have further thoughts.}
It can consist of fundamental types, such as integers or booleans.
There may also be parameterised or composite types, for example lists may have a type parameter defining what values it can contain.
This enables differentiating lists that store integers from those that store strings and ensuring that any item added to a list must be of the same type as the type parameter.
%\CR{A bit vague; consider giving an example.}
\begin{figure}[h]
\begin{lstlisting}[]
abstract Type;

Bool : Type;
Int  : Type;
\end{lstlisting}
  \caption{Type syntax consisting of two types, for boolean and integer values, written as a JastAdd AST specification (see Section \ref{astdef})}
  \label{typesyntaxspec}
\end{figure}

@@

A slightly more complex case can be expressed as \lstinline{if true then true else true}.
The only rule that matches the structure, is T-If, with \lstinline{true} values taking the place of each of the $t_i$ variables.
To evaluate this rule, we are thus required to recursively evaluate the types each of our $t_i$ variables, to ensure they match the premises.
In this case, the type of $t_1$ evaluates to \lstinline{Bool}, matching the first premise of T-If.
The types of both $t_2$ and $t_3$, represented as the type variable \tyvar{T}, also evaluate to \lstinline{Bool}.
Since both the uses of this type variable within the premises are consistent, all three of the rule's premises have been fulfilled, and the entire expression can be typed as \tyvar{T}, which has been found to correspond to \lstinline{Bool}.

@@

\subsection{JastAdd}
JastAdd is an implementation of RAGs for Java, offering a flexible system for writing language analysis in an object-oriented fashion.\cite{JastAdd}
Our project uses it both for the implementation of our compiler and as a compilation target, as it outputs JastAdd code.
%\CR{It's worth telling the reader here why you are explaining some details about JastAdd.}

\subsubsection{AST specification}\label{astdef}
One of the key components of JastAdd projects are the \lstinline{.ast} files, which declare the structure of the abstract syntax tree.
This specification is then used by JastAdd to generate Java classes representing the nodes of the AST.
These AST classes can have a variety of forms, supporting both abstract classes and subclasses.
They can also contain components such as Java types, tokens and child nodes, and these components may also be optional or list components.
For a detailed overview of the syntax, see the reference manual\cite{jastaddref}.

\subsubsection{RAG Modules}
\begin{figure}[h]
\begin{lstlisting}[language=jrag]
aspect TypingRules {
  syn Type Zero.type() {
    return new Int();
  }
}
\end{lstlisting}
  \caption{Snippet of a .jrag file, declaring a type attribute for the Zero AST node}
  \label{jragexample}
\end{figure}
The RAG modules are written in \lstinline{.jrag} files and declare the attributes to be included in the classes generated by the AST specification.
Figure \ref{jragexample} shows a minimal example.
The files use a syntax very similar to that of Java to declare attributes of a variety of different kinds, and the body of methods corresponds directly to Java code.

\chapter{Approach/Design} % should contain a description of your solution(s), with all the theoretical background needed. On occasion this is replaced by a subset or all of the following:

@@

uppercase_id = "[A-Z][A-Za-z0-9-]";
uppercase_id = "[a-z][A-Za-z0-9-]";
\end{lstlisting}
  \caption{A definition of the concrete syntax for the typing rule language in a notation similar to BNF.}
  \label{trconcretesyntaxspec}
\end{figure}
%\CR{After reading, I would now say that it's preferable to me if you keep the language definition separate
%  from its implementation.  You can use Figure 3.4 as the abstract syntax, to complement the current discussion.
%  This would be before the current 3.1, then.
%  An advantage of this is that you can then start your ``Architecture'' discussion (in 3.2) by mentioning
%  the complex dependencies, i.e., that processing a type system specification requires you to know the abstract syntax
%  of the object language as well as of the types.}

@@

To express typing rules, we have defined an ASCII representation of the natural semantics syntax used by Pierce\cite{Pierce}.
Our language can express a subset of the notation used by Pierce, with some differences.

@@

\subsubsection{Typing rules parser}\label{typingrulesparser}
The parser for typing rules is written using JFlex\cite{jflex}, a scanner generator, and Beaver\cite{beaver}, an LALR parser generator.
JFlex takes a series of regular expressions to generate a tokenizer for the parsed language.
The tokens are passed on to Beaver along with a parser definition in extended Backus-Naur form (EBNF), to generate a parser.

\subsubsection{Typing rules AST}\label{typingrulesast}
%\CR{I would give this AST/BNF a more specific name here, to distinguish it from the other two (as done nicely already in Figure 3.1)}
\begin{figure}[h]
\begin{lstlisting}[]
RuleSet ::= Rule*;

@@

\subsection{Support library}
In addition to the generated type-checking files, there are a number of files that are added to the output file that are identical for all generated type checkers, which we refer to as the support library.
Some of these are key components of the type checker, but others are dummy components included solely for convenience of testing.
We expect that in a real world use case, our software would integrate into an object language's compiler pipeline, and utilise its own parser and project files to provide the abstract syntax tree for our type checking to tie into.
For testing purposes, we have written generic versions of these components, which work for any supported object language.
%\CR{``Templates'' are usually things that are mostly fixed but have holes to be filled in.  I would call what you describe here the ``support library''}

@@

This approach does have the issue of occasionally parsing invalid programs, due to the reflection API finding an additional parameterless default constructor.
This leads the parser to always accept nodes written with no children, such as \lstinline{Expression(Or())}, even if the AST specification requires them.
These children will be null, causing issues further down the line, though as this parser is only intended for testing it is of little concern.
%\CR{Does this also apply to the analysis of typing rules?  If so, you can mention it as ``future work''
%  or describe how it could be fixed (it's not a fundamental limitation, so nothing that warrants a deeper discussion)}

\subsubsection{Project files}\label{projectfiles}
For ease of use the compiler will output a complete runnable Gradle project.
It can parse simplified tree-representations of programs, and utilise these to perform the type checking.
In a future real world use case, it is intended for our generated type-checking aspects to integrate into a pre-existing project, which already has all the necessary project files to parse an AST tree, to which our type aspects could integrate.

@@

\subsection{Type-checking algorithm}\label{typecheckingalgorithm}

@@

\begin{figure}[]
%foreach parameter in rule.conclusion.parameters:
%  declare var "parameer.name" = getChild(parameter.index)
\begin{lstlisting}[]
foreach typevar in rule.typevars:
  typevar.type = null;
  foreach premise in rule.premises where premise.rhs == typevar:
    if typevar.type == null:
      typevar.type = premise.lhs.type()
    else if typevar.type != premise.lhs.type():
      throw typeerror("Type variable mismatch")

@@

\subsubsection{Aspect generation}
The \lstinline{RuleSet} node is responsible for generating the overall structure of the output file, a JastAdd aspect called \lstinline{TypingRules}.
It sorts its \lstinline{Rule} children by which nodes they apply to, so that if multiple tying rules type the same node, they can be inserted into the same method implementation (though this feature is not yet fully working).

@@

\subsubsection{Parameters}
When the typing rules concern a node with parameters, these can be named arbitrarily in the typing rule.
%\CR{Who does the naming?  Why is the ``arbitrarily'' bit important?}
%\CR{ Will this work if the type system specification has a type variable called ``this'' or ``null'' or ``class''?}.
A variable with the same name will be instantiated at the beginning of the method, fetching the corresponding child node by index.
This naming format means any errors reported will use the same variable names as the user has defined in the typing rules, in order to minimise the disconnect when analysing errors from the generated code.
However, it also means it is not possible to name a parameter any of the reserved keywords in Java.
This could be alleviated by adding a prefix to the variable names, at the cost of a slight indirection for the user.
These variables are then utilised in later premises or type variable declarations.

\subsubsection{Type variables}
In the case of type variables, much of the logic is handled at runtime, to simplify the code generation.
Each type variable that appears in the conclusion is instantiated to a null value, and the code generated by each premise first checks if the value is still null and either assigns the variable its own type or tests to make sure the existing type matches its own.

@@

\begin{figure}[]
\begin{lstlisting}[]
[T-If]
x: Bool,
y: a,
z: a
------------------
If(x, y, z): a
\end{lstlisting}
  \caption{Additional typing rules for the BoolsIf language, an extension for the Bools typing rules in Figure \ref{boolstr} with an if term}
  \label{iftr}
\end{figure}
\begin{figure}[]
\begin{lstlisting}[language=jrag]
syn Type If.type() {
  ASTNode x = getChild(0);
  ASTNode y = getChild(1);
  ASTNode z = getChild(2);

@@

%\CR{``Shown in Figure~\ref{reference}'' is the current preferred form in our field}
The code for this rule is shown in Figure \ref{iftrgen}, and begins with three parameter declarations as in previous rules, followed by a new type variable declaration on line 6, which is instantiated to null.
For each invocation of the type variable, a code section is generated, which checks if the type variable is still null, and if so assigns it to the type of the left-hand side term, or if it has been assigned checks whether the types match each other. Should any non-matches be found, a type error is thrown.

The rest of the method is similar, with premises being checked -- including a redundant check of the type variables -- before returning the type variable, or a type error being thrown.

@@

The current implementation of \lstinline[language=java]{matches()} would however consider all lists to match, without comparing the child nodes.
Perhaps the default implementation could be extended to compare all child elements, at the risk of potentially tricky evaluation, or these more advanced types would require manually overriding the method.
Yet another approach may be to design a way to define type equality within the typing rules, allowing users to customise the behaviour without having to restort to manually writing JastAdd code.

@@

We also ran into a few tooling issues.
The first was Gradle not willing to recognise JastAdd as both a library and a build component, which we eventually solved by manually providing a jar file separately from the resolved build dependency.
The second one was our LSP server, jdtls, not recognising JastAdd imports, which we never managed to fix.
%Gradle troubles
%Library not recognised by LSP server
%Difficult runtime parsing of AST files

@@

\subsection{Handle multiple rules for the same node}
In type systems it is generally considered valid to have multiple typing rules for the same construct, with different, non-intersecting prerequisites.
It could for example be useful to type an addition operator for a variety of different numerical types, which may not have a common ancestor.

@@

\vspace*{\fill}
\hspace*{\fill}
\includegraphics[scale=0.2]{stagbeetle.pdf}

@@

%TODO: Re-add the popsci summary
%%make sure we're on even page with the pop-sci
%\checkoddpage
%\ifoddpage
%\else
%  \newpage
%  \thispagestyle{empty}
%  \mbox{ }
%\fi
%\includepdf[pages={1}]{popsci/popsci.pdf}
\end{appendices}

>>>>

\thesisnumber{LU-CS-EX: 2024-06} % Magic Number! Do not change unless Birger Swahn asks you to do so!
% default is Master. Uncomment the following for "kandidatarbete"/Bachelor's thesis
%\thesistype{Bachelor}{Kandidatarbete}

@@

%\title{Formatting a Master's Thesis}
\title{Type Checker Generation using Reference Attribute Grammars}
\svensktitel{Generering av Typkontrollerare med Referens-Attribut-Grammatik}

@@

\acknowledgements{
Sincere thanks to Christoph Reichenbach, whose enthusiasm for the field inspired me to pursue this project and whose guidance in matters of both implementation and writing was key to this thesis.
Additional thanks to Niklas Fors for providing several valuable insights, greatly simplifying certain components and Alexandru Dura for providing some guidance to JastAdd internals.

I would also like to thank my partner Lo-Isabel Krantz Andrée as well as my family, friends, and Moritz for their support throughout the duration of this project.

%If you want to thank people, do it here, on a separate right-hand page. Both the U.S. \textit{acknowledgments} and the British \textit{acknowledgements} spellings are acceptable.
%
%We would like to thank Lennart Andersson for his feedback on this template.
%
%We would also like thank Camilla Lekebjer for her contribution on this template, as well as Magnus Hultin for his popular science summary class and example document.
%
%Thanks also go to the following (former) students for helping with feedback and suggestions on this template: Mikael Persson, Christoffer Lundgren, Mahmoud Nasser.
\vspace*{\fill}
\hspace*{\fill}
\includegraphics[scale=0.5]{assets/cat.pdf}
}

@@

We define a custom parsable language for typing rules adapted from natural semantics, and introduce a strategy to compile it into RAG attributes.
Our case studies show that simple type systems can be systematically compiled into attributes on the Abstract Syntax Tree (AST), and the tree syntax provides a natural approach for recursive evaluation.
Overall, this initial exploration shows that RAGs are a flexible approach for implementing type checkers for syntax-directed typing rules, though there are difficulties when implementing rule verification at compiler runtime.
%\CR{Usually you then put a high-level summary of the rest of the thesis here (not mentioning chapters yet), something like ``we examine X and Y and introduce a strategy for Z.
%  We evaluate the approach against Foo, and demonstrate Bar.
%  Our case studies show Quux.
%  Overall, this initial exploration shows \ldots''}

@@

\section{Abstract syntax trees}\label{abstractsyntaxtrees}
\begin{figure}[]
  \centering
  \begin{tikzpicture}[
      every node/.style = {align=center, line width=0.8pt},
      every path/.style = {draw, -latex},
      astnode/.style={draw=black},
      %greennode/.style={shape=circle, draw=green, line width=2},
      %rednode/.style={shape=circle, draw=red, line width=2}
    ]
    \node (if)   [astnode]                             {If};
    \node (then) [astnode, below=of if]                {True};
    \node (else) [astnode, right=of then]              {False};
    \node (cond) [astnode, left=of then]               {GreaterThan};
    \node (one)  [astnode, below = of cond.south west] {Int\\ \textcolor{gray}{value: 1}};
    \node (zero) [astnode, below = of cond.south east] {Int\\ \textcolor{gray}{value: 0}};

    %\node () [above right = 0.3cm and -1cm of cond] {cond};
    %\node () [above left = 0.1cm and -0.7cm of then] {then};
    %\node () [above left= 0.3cm and -0.5cm of else] {else};
    %\node () [above= 0.3cm of one] {left};
    %\node () [above= 0.3cm of zero] {right};

    \draw (if.south)   to (cond.north);
    \draw (if.south)   to (then.north);
    \draw (if.south)   to (else.north);
    \draw (cond.south) to (one.north);
    \draw (cond.south) to (zero.north);
    %\draw (if.south)   to [edge label=Cond]  (cond.north);
    %\draw (if.south)   to [edge label=Then]  (then.north);
    %\draw (if.south)   to [edge label=Else]  (else.north);
    %\draw (cond.south) to [edge label=Left]  (one.north);
    %\draw (cond.south) to [edge label=Right] (zero.north);

  \end{tikzpicture}
  \caption{Example of an abstract syntax tree}
  \label{astexample}
\end{figure}
The syntax of a language is the definition of the symbols and structures which make up a valid program.
%\CR{What's a syntax?}
An abstract syntax tree (AST) is a simplified tree-based representation of the syntax.
Each node in the tree represents a syntactic construct of the language, such as expressions, statements or declarations and the branches represent the structure.
It is abstract in that it neglects to include all the concrete details of the syntax, but rather focuses on the structure of the program.
Structural elements of the source code, such as parenthesis, commas and brackets, are no longer necessary, as the structure of the tree conveys this semantic information.

This makes it easier to write analysis based on the behaviour of a program, without concerning oneself with its exact formulation.
Several different formulations of a program with identical behaviour may be abstracted into identical ASTs.
The hierarchical structure of the tree lends itself especially well for recursive analyses, and eases the passing of information between connected structures by connecting them as parent and children nodes.
%\CR{what is ``this information'', concretely?}
Figure \ref{astexample} shows an example of an abstract syntax tree representation of the expression \verb|if 1 > 0 then true else false|.
%A statement like \verb|if (true) then {1} else {0}| may instead be represented by an if node, with named branches pointing towards the condition and the two values.
%\CR{You could show an example, but at the very least I would like to read an argument for why the hierarchical structure of a tree is helpful.}
%\CR{You can find common examples for showing the value of ASTs over concrete syntax trees in the EDAN65 slides.}

@@

\subsection{Type checking}
Verifying that these constraints are upheld can assist us in catching programming mistakes, preventing the wrong data types from being used in the wrong places, which may have led to erroneous or unpredictable behaviour.
There are two main approaches to ensuring the constraints are upheld, dynamic and static type checking.
As our project generates a static type checker, we will focus our explanations on that topic, but provide a quick overview of their differences here.
%\CR{Why would we even want to ensure that they are upheld?}

@@

\subsubsection{Type syntax}\label{typesyntax}
We're using the term `type syntax' to mean a definition of the valid types within a language, often referred to simply as the set of types.
%\CR{Is ``type syntax'' what Pierce calls it?  Not sure I've heard that term before; I don't think it's wrong, personally, but please check in case he uses a different term.}
%\reviewquestion{I believe Pierce simply refers to it as the "set of types". I seem to have got the term type syntax from EDAP15 \url{https://fileadmin.cs.lth.se/cs/Education/EDAP15/2022/web/skills/type-system.html}. In my opinion it works better to have a clear name for it, so I'm keeping it for now, unless you have further thoughts.}
It can consist of fundamental types, such as integers or booleans.
There may also be parameterised or composite types, for example lists may have a type parameter defining what values it can contain.
This enables differentiating lists that store integers from those that store strings and ensuring that any item added to a list must be of the same type as the type parameter.

Figure \ref{typesyntaxspec} shows an example specification of a type syntax, consisting of only two types, integers and booleans.
This specification is used for all the languages used in our project.
%\CR{A bit vague; consider giving an example.}
\begin{figure}[]
\begin{lstlisting}[]
abstract Type;

Bool : Type;
Int  : Type;
\end{lstlisting}
  \caption{Type syntax consisting of two types, for boolean and integer values, written as a JastAdd AST specification (see Section \ref{astdef}).}
  \label{typesyntaxspec}
\end{figure}

@@

A somewhat more complex case can be expressed as \lstinline{if true then true else }\\\lstinline{true}.
The only rule that matches the structure, is T-If, with \lstinline{true} values taking the place of each of the $t_i$ variables.
To evaluate this rule, we are thus required to recursively evaluate the types each of our $t_i$ variables, to ensure they match the premises.
In this case, the type of $t_1$ evaluates to \lstinline{Bool}, matching the first premise of T-If.
The types of both $t_2$ and $t_3$, represented as the type variable \tyvar{T}, also evaluate to \lstinline{Bool}.
Since both the uses of this type variable within the premises are consistent, all three of the rule's premises have been fulfilled, and the entire expression can be typed as \tyvar{T}, which has been found to correspond to \lstinline{Bool}.

@@

\subsection{JastAdd}
JastAdd is an implementation of RAGs for Java, offering a flexible system for writing language analysis in an object-oriented fashion\cite{JastAdd}.
Our project uses it both for the implementation of our compiler and as a compilation target, as it outputs JastAdd code.
%\CR{It's worth telling the reader here why you are explaining some details about JastAdd.}

\subsubsection{AST specification}\label{astdef}
\begin{figure}[]
\begin{lstlisting}[]
Expression ::= Term;

abstract Term;
True  : Term;
False : Term;
Or    : Term ::= Left:Term Right:Term;
Zero  : Term;

\end{lstlisting}
  \caption{AST specification for the Bools language.}
  \label{astspecexample}
\end{figure}
One of the key components of JastAdd projects are the \lstinline{.ast} files, which declare the structure of the abstract syntax tree.
This specification is then used by JastAdd to generate Java classes representing the nodes of the AST.
These AST classes can have a variety of forms, supporting both abstract classes and subclasses.
They can also contain components such as Java types, tokens and child nodes, and these components may also be optional or list components.

Figure \ref{astspecexample} shows a small example of the specificaton syntax.
Each of the node definitions begins with a class name.
\lstinline{True : Term;} indicates that the \lstinline{True} node is a subclass of the abstract class \lstinline{Term}.
If a node has child nodes, these are provided after the \lstinline{::=}, and may either have a name and a type separated by a colon, or only a type.
For a detailed overview of the syntax, see the reference manual\cite{jastaddref}.

\subsubsection{RAG Modules}
\begin{figure}[]
\begin{lstlisting}[language=jrag]
aspect TypingRules {
  syn Type Zero.type() {
    return new Int();
  }
}
\end{lstlisting}
  \caption{Snippet of a .jrag file, declaring a type attribute for the Zero AST node}
  \label{jragexample}
\end{figure}
The RAG modules are written in \lstinline{.jrag} files and declare the attributes to be included in the classes generated by the AST specification.
Figure \ref{jragexample} shows a minimal example.
The files use a syntax very similar to that of Java to declare attributes of a variety of different kinds, and the body of methods corresponds directly to Java code.

\chapter{Approach} % should contain a description of your solution(s), with all the theoretical background needed. On occasion this is replaced by a subset or all of the following:

@@

uppercase_id = "[A-Z][A-Za-z0-9-]";
uppercase_id = "[a-z][A-Za-z0-9-]";
\end{lstlisting}
  \caption{A definition of the concrete syntax for the typing rule language in a notation similar to Backus-Naur form (BNF).}
  \label{trconcretesyntaxspec}
\end{figure}
%\CR{After reading, I would now say that it's preferable to me if you keep the language definition separate
%  from its implementation.  You can use Figure 3.4 as the abstract syntax, to complement the current discussion.
%  This would be before the current 3.1, then.
%  An advantage of this is that you can then start your ``Architecture'' discussion (in 3.2) by mentioning
%  the complex dependencies, i.e., that processing a type system specification requires you to know the abstract syntax
%  of the object language as well as of the types.}

@@

To express typing rules, we have defined an ASCII representation of the natural semantics syntax used by Pierce\cite{Pierce}, with its concrete syntax given in Figure \ref{trconcretesyntaxspec}.
Our language can express a subset of the notation used by Pierce, with some differences.

@@

\subsubsection{Typing rules parser}\label{typingrulesparser}
The parser for typing rules is written using JFlex\cite{jflex}, a scanner generator, and Beaver\cite{beaver}, a parser generator.
JFlex takes a series of regular expressions to generate a tokenizer for the parsed language.
The tokens are passed on to Beaver along with a parser definition in extended Backus-Naur form (EBNF), to generate a parser.

\subsubsection{Typing rules AST}\label{typingrulesast}
%\CR{I would give this AST/BNF a more specific name here, to distinguish it from the other two (as done nicely already in Figure 3.1)}
\begin{figure}[]
\begin{lstlisting}[]
RuleSet ::= Rule*;

@@

\subsection{Support library}
In addition to the generated type checking files, there are a number of files that are added to the output file that are identical for all generated type checkers, which we refer to as the support library.
Some of these are key components of the type checker, but others are dummy components included solely for convenience of testing.
We expect that in a real world use case, our software would integrate into an object language's compiler pipeline, and utilise its own parser and project files to provide the abstract syntax tree for our type checking to tie into.
For testing purposes, we have written generic versions of these components, which work for any supported object language.
%\CR{``Templates'' are usually things that are mostly fixed but have holes to be filled in.  I would call what you describe here the ``support library''}

@@

This approach does have the issue of occasionally parsing invalid programs, due to the reflection API finding an additional parameterless default constructor.
This leads the parser to always accept nodes written with no children, such as \lstinline{Expression(Or())}, even if the AST specification requires them.
These children will be null, causing issues further down the line, though as this parser is only intended for testing it is of little concern.
Our tests have been carefully written to avoid this error, and any faulty tests would result in obvious null pointer exceptions during further evaluation.
%\CR{Does this also apply to the analysis of typing rules?  If so, you can mention it as ``future work''
%  or describe how it could be fixed (it's not a fundamental limitation, so nothing that warrants a deeper discussion)}

\subsubsection{Project files}\label{projectfiles}
For ease of use the compiler will output a complete runnable Gradle project.
It can parse simplified tree-representations of programs, and utilise these to perform the type checking.
In a future real world use case, it is intended for our generated type checking aspects to integrate into a pre-existing project, which already has all the necessary project files to parse an AST tree, to which our type aspects could integrate.

@@

\subsection{Type checking algorithm}\label{typecheckingalgorithm}

@@

\begin{figure}[bhpt]
%foreach parameter in rule.conclusion.parameters:
%  declare var "parameer.name" = getChild(parameter.index)
\begin{lstlisting}[]
foreach typevar in rule.typevars:
  typevar.type = null;
  foreach premise in rule.premises where premise.rhs == typevar:
    if typevar.type == null:
      typevar.type = premise.lhs.type()
    else if typevar.type != premise.lhs.type():
      throw typeerror("Type variable mismatch")

@@

\subsubsection{Aspect generation}
The \lstinline{RuleSet} node is responsible for generating the overall structure of the output file, a JastAdd aspect called \lstinline{TypingRules}.
It sorts its \lstinline{Rule} children by which nodes they apply to, so that if multiple typing rules type the same node, they can be inserted into the same method implementation (though this feature is not yet fully working).

@@

\subsubsection{Child nodes}
When the typing rules concern a node with children, these are bound to free variables which can be named arbitrarily.
%\CR{Who does the naming?  Why is the ``arbitrarily'' bit important?}
%\CR{ Will this work if the type system specification has a type variable called ``this'' or ``null'' or ``class''?}.
A variable with the same name will be instantiated at the beginning of the method, fetching the corresponding child node by index.
This naming format means any errors reported will use the same variable names as the user has defined in the typing rules, in order to minimise the disconnect when analysing errors from the generated code.
However, it also means it is not possible to name a variable any of the reserved keywords in Java.
This could be alleviated by adding a prefix to the variable names, at the cost of a slight indirection for the user.
These variables are then utilised in later premises or type variable declarations.

\subsubsection{Type variables}\label{typevariables}
In the case of type variables, much of the logic is handled at runtime, to simplify the code generation.
Each type variable that appears in the conclusion is instantiated to a null value, and the code generated by each premise first checks if the value is still null and either assigns the variable its own type or tests to make sure the existing type matches its own.

@@

\begin{figure}[]
\begin{lstlisting}[]
[T-If]
x: Bool,
y: a,
z: a
--------------
If(x, y, z): a
\end{lstlisting}
  \caption{Additional typing rules for the BoolsIf language, an extension for the Bools typing rules in Figure \ref{boolstr} with an if term}
  \label{iftr}
\end{figure}
\begin{figure}[]
\begin{lstlisting}[language=jrag]
syn Type If.type() {
  ASTNode x = getChild(0);
  ASTNode y = getChild(1);
  ASTNode z = getChild(2);

@@

%\CR{``Shown in Figure~\ref{reference}'' is the current preferred form in our field}
The code for this rule is shown in Figure \ref{iftrgen}, and begins with three local variable declarations as in previous rules, followed by a new type variable declaration on line 6, which is instantiated to null.
For each invocation of the type variable, a code section is generated, which checks if the type variable is still null, and if so assigns it to the type of the left-hand side term, or if it has been assigned checks whether the types match each other. Should any non-matches be found, a type error is thrown.

The rest of the method is similar, with premises being checked -- including a redundant check of the type variables, as mentioned in Section \ref{typevariables} -- before returning the type variable, or a type error being thrown.
This type checker accurately captures the intended behaviour of the typing rule.

@@

The current implementation of \lstinline[language=java]{matches()} would however consider all lists to match, without comparing the child nodes.
Perhaps the default implementation could be extended to compare all child elements, at the risk of potentially tricky evaluation, or these more advanced types would require manually overriding the method.
Yet another approach may be to design a way to define type equality within the typing rules, allowing users to customise the behaviour without having to resort to manually writing JastAdd code.

@@

We also ran into a few tooling issues.
The first was Gradle not willing to recognise JastAdd as both a library and a build component, which we eventually solved by manually providing a jar file separately from the resolved build dependency.
The second one was our Language Server Protocol (LSP) server, jdtls, not recognising JastAdd imports and flagging every use thereof as an error, which we never managed to fix.
%Gradle troubles
%Library not recognised by LSP server
%Difficult runtime parsing of AST files

@@

\subsection{Handle multiple rules for the same node}
In type systems it is generally considered valid to have multiple typing rules for the same construct, with different, non-intersecting prerequisites.
It could, for example, be useful to type an addition operator for a variety of different numerical types, which may not have a common ancestor.

@@

\vspace*{\fill}
\hspace*{\fill}
\includegraphics[scale=0.2]{assets/stagbeetle.pdf}

@@

%make sure we're on even page with the pop-sci
\checkoddpage
\ifoddpage
\else
  \newpage
  \thispagestyle{empty}
  \mbox{ }
\fi
\includepdf[pages={1}]{popsci/popsci.pdf}
\end{appendices}
